---
title: "QTP"
author: "Alex Hirtle - 1719121"
editor: visual
format:
  html:
    embed-resources: true
    self-contained: true
    number-sections: true
    toc: true
    toc-location: left
    toc-depth: 2
    toc-title: "Contents"
---

```{r, "Chunk Opts", echo = F, warning = F, message = F}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r, "Libraries", include = F}
library(tidyverse)
library(tidyquant)
library(PerformanceAnalytics)
library(cluster)
library(gt)
library(slider)
library(doParallel)
library(dlm)
library(purrr)
library(patchwork)
library(timetk)
library(plotly)
```

```{r, "Pulling Data", include = F}
tickers <- c("SPY", "XLB", "XLC", "XLE",
             "XLF", "XLI", "XLK", "XLP",
             "XLRE", "XLV", "XLU", "XLY")

fromdate <- "2000-06-01"
todate   <- "2026-02-28"

raw_data   <- tq_get(tickers, get = "stock.prices", from = fromdate, to = todate) %>%
  stats::na.omit()
train_data <- raw_data %>% dplyr::filter(date <= "2024-12-31")
test_data  <- raw_data %>% dplyr::filter(date >= "2025-01-01")
```

```{r, "Returns", include = F}
# Adds OpCl, ClCl, ClOp log returns to the df per symbol.
add_returns <- function(dat) {
  dat %>%
    dplyr::group_by(symbol) %>%
    dplyr::mutate(
      OpCl = (close    / open)                    - 1,  # intraday: same-day prices, no adj needed
      ClCl = (adjusted / dplyr::lag(adjusted))    - 1,  # total daily return including dividends
      ClOp = (open     / dplyr::lag(adjusted))    - 1   # overnight: prev adj close to today's open
    ) %>%
    stats::na.omit() %>%
    dplyr::ungroup()
}

train_df <- add_returns(train_data)
test_df  <- add_returns(test_data)

write.csv(test_df, "Testing_DF_Returns")
```

```{r, "Rolling Stats", include = F}
# Adds rolling mean (short + long for MA crossover), vol, vol-of-vol,
# volume mean, Sharpe, momentum, SPY correlation, and max drawdown.
#
# When `window` is supplied it acts as a global override: all stats use that
# window, and mean_windows is automatically set to c(round(window/4), window)
# to produce both the short and long MA required for the crossover signal.
# Individual window arguments are ignored when `window` is set.
#
# Args:
#   dat              : Grouped or ungrouped df with symbol, date, ClCl,
#                      volume columns. Must contain "SPY" for corr stats.
#   window           : Global window override (integer). Short MA = round(window/4).
#   mean_windows     : Windows for rolling mean of ClCl. Ignored if window set.
#   vol_windows      : Windows for rolling vol (sd of ClCl).
#   vol_of_vol_windows: Outer windows for rolling sd of the base vol series.
#   vol_base_window  : Inner window for VoV base. Defaults to vol_windows[1].
#   volume_windows   : Windows for rolling mean of share volume.
#   sharpe_windows   : Windows for rolling annualised Sharpe.
#   momentum_windows : Windows for cumulative compound return.
#   corr_windows     : Windows for rolling Pearson correlation to SPY.
#   drawdown_windows : Windows for rolling max drawdown.
#   rf_daily         : Daily risk-free rate for Sharpe. Default 0.
#
# Returns:
#   df with all rolling columns appended, ungrouped.
#   Rows without a full window are NA — no look-forward bias.

add_rollers <- function(
  dat,
  window              = NULL,
  mean_windows        = c(20, 252),
  vol_windows         = c(20, 60),
  vol_of_vol_windows  = c(60),
  vol_base_window     = NULL,
  volume_windows      = c(20),
  sharpe_windows      = c(20, 60),
  momentum_windows    = c(20, 60, 252),
  corr_windows        = c(60),
  drawdown_windows    = c(60),
  rf_daily            = 0
) {

  # If global window supplied, override all individual window params.
  # Short MA = window / 4 (1:4 ratio) for the crossover signal.
  if (!is.null(window)) {
    short_w            <- max(2L, round(window / 4L))
    mean_windows       <- c(short_w, window)
    vol_windows        <- window
    vol_of_vol_windows <- window
    volume_windows     <- window
    sharpe_windows     <- window
    momentum_windows   <- window
    corr_windows       <- window
    drawdown_windows   <- window
  }

  # ---------------------------------------------------------------------------
  # Internal helpers
  # Args: x = series, w = window, f = function, y = second series (corr only)
  # ---------------------------------------------------------------------------

  # Right-aligned rolling apply; NA until full window is available.
  roll_scalar <- function(x, w, f) {
    slider::slide_dbl(x, f, .before = w - 1, .complete = TRUE)
  }

  # Rolling Pearson correlation between two parallel vectors.
  roll_corr <- function(x, y, w) {
    slider::slide2_dbl(x, y,
      ~ cor(.x, .y, use = "complete.obs"),
      .before = w - 1, .complete = TRUE)
  }

  # Annualised Sharpe (252-day). NA when sd = 0.
  sharpe_fn <- function(x, rf) {
    excess <- x - rf
    s <- sd(excess, na.rm = TRUE)
    if (is.na(s) || s == 0) return(NA_real_)
    sqrt(252) * mean(excess, na.rm = TRUE) / s
  }

  # Max drawdown over the window as a negative fraction.
  mdd_fn <- function(x) {
    cr   <- cumprod(1 + x)
    peak <- cummax(cr)
    min((cr - peak) / peak, na.rm = TRUE)
  }

  # Cumulative compound return.
  momentum_fn <- function(x) prod(1 + x, na.rm = TRUE) - 1

  # ---------------------------------------------------------------------------
  # Prepare: join SPY ClCl for correlation, group by symbol, sort by date.
  # ---------------------------------------------------------------------------
  spy_series <- dat %>%
    dplyr::filter(symbol == "SPY") %>%
    dplyr::arrange(date) %>%
    dplyr::select(date, spy_ClCl = ClCl)

  dat <- dat %>%
    dplyr::left_join(spy_series, by = "date") %>%
    dplyr::group_by(symbol) %>%
    dplyr::arrange(date, .by_group = TRUE)

  # Rolling means (produces mean_<short>d and mean_<long>d when window set)
  for (w in mean_windows) {
    nm  <- paste0("mean_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(ClCl, w, ~ mean(.x, na.rm = TRUE)))
  }

  # Rolling volatility (sd of ClCl)
  for (w in vol_windows) {
    nm  <- paste0("vol_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(ClCl, w, ~ sd(.x, na.rm = TRUE)))
  }

  # Volatility of volatility: rolling sd of the base vol series
  base_w       <- if (!is.null(vol_base_window)) vol_base_window else vol_windows[1]
  base_vol_col <- paste0("vol_", base_w, "d")

  for (w in vol_of_vol_windows) {
    nm  <- paste0("vov_", base_w, "d_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(.data[[base_vol_col]], w, ~ sd(.x, na.rm = TRUE)))
  }

  # Rolling mean of share volume
  for (w in volume_windows) {
    nm  <- paste0("volume_mean_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(volume, w, ~ mean(.x, na.rm = TRUE)))
  }

  # Rolling Sharpe ratio (annualised)
  for (w in sharpe_windows) {
    nm  <- paste0("sharpe_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(ClCl, w, ~ sharpe_fn(.x, rf_daily)))
  }

  # Rolling momentum (cumulative compound return)
  for (w in momentum_windows) {
    nm  <- paste0("momentum_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(ClCl, w, momentum_fn))
  }

  # Rolling correlation to SPY
  for (w in corr_windows) {
    nm  <- paste0("corr_spy_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_corr(ClCl, spy_ClCl, w))
  }

  # Rolling max drawdown
  for (w in drawdown_windows) {
    nm  <- paste0("mdd_", w, "d")
    dat <- dat %>%
      dplyr::mutate(!!nm := roll_scalar(ClCl, w, mdd_fn))
  }

  dat %>%
    dplyr::select(-spy_ClCl) %>%
    dplyr::ungroup()
}
```

```{r, "Kalman Smoothing", include = F}
# Applies a local-level Kalman filter to every rolling stat column.
# Each series is normalised to zero mean / unit variance before filtering
# so that the same dV and dW are meaningful across all columns regardless
# of scale differences (e.g. returns vs Sharpe ratios vs volume).
#
# Local level model:
#   mu_t = mu_{t-1} + w_t,   w_t ~ N(0, dW)   state evolves as random walk
#   y_t  = mu_t    + v_t,   v_t ~ N(0, dV)   observation = state + noise
#
# The ratio dW/dV controls smoothing:
#   Low  dW/dV → state changes slowly → heavy smoothing
#   High dW/dV → state changes quickly → light smoothing, tracks data closely
#
# dV is fixed at 1 (only the ratio matters after normalisation).
# The filter is strictly causal — no look-forward bias.
#
# Args:
#   dat : df output from add_rollers. Must have symbol and date columns.
#   dW  : State noise variance. The key sensitivity-analysis parameter.
#   dV  : Observation noise variance. Fixed at 1; series are pre-normalised.
#
# Returns:
#   dat with rolling stat columns replaced by their Kalman-filtered versions.

kalman_smooth_all <- function(dat, dW, dV = 1) {

  # Columns that should NOT be filtered (raw prices, returns, identifiers)
  raw_cols <- c("symbol", "date", "open", "high", "low", "close",
                "volume", "adjusted", "OpCl", "ClCl", "ClOp")

  feature_cols <- setdiff(names(dat), raw_cols)
  feature_cols <- feature_cols[vapply(dat[feature_cols], is.numeric, logical(1))]

  local_level <- dlm::dlmModPoly(order = 1, dV = dV, dW = dW)

  # Filter a single numeric vector: normalise → Kalman filter → denormalise.
  filter_series <- function(x) {
    valid <- which(is.finite(x))
    if (length(valid) < 10) return(x)  # too few points to filter meaningfully

    x_sub  <- x[valid]
    x_mean <- mean(x_sub)
    x_sd   <- sd(x_sub)

    if (is.na(x_sd) || x_sd == 0) return(x)  # constant series; nothing to smooth

    x_norm   <- (x_sub - x_mean) / x_sd
    filtered <- tryCatch(
      dlm::dlmFilter(x_norm, local_level),
      error = function(e) NULL
    )
    if (is.null(filtered)) return(x)

    # filtered$m is T+1 rows; row 1 = prior. Rows 2:(T+1) = filtered states.
    x_smooth     <- as.numeric(filtered$m[-1])
    result       <- x
    result[valid] <- x_smooth * x_sd + x_mean
    result
  }

  dat %>%
    dplyr::group_by(symbol) %>%
    dplyr::arrange(date, .by_group = TRUE) %>%
    dplyr::mutate(dplyr::across(dplyr::all_of(feature_cols), filter_series)) %>%
    dplyr::ungroup()
}
```

```{r, "Monthly Snapshots", include = F}
# Extracts the last trading day of each calendar month per symbol.
# These are the rebalancing observation points — all clustering and signal
# logic operates on this reduced dataframe.
#
# Args:
#   dat : Daily returns df with symbol and date columns.
#
# Returns:
#   df with one row per symbol per month-end date.

get_monthly_snapshots <- function(dat) {
  dat %>%
    dplyr::mutate(.ym = format(date, "%Y-%m")) %>%
    dplyr::group_by(symbol, .ym) %>%
    dplyr::filter(date == max(date)) %>%
    dplyr::ungroup() %>%
    dplyr::select(-.ym)
}
```

```{r, "Clustering", include = F}
# evaluate_k: compares cluster quality across multiple K values using
# within-cluster sum of squares (elbow method) and average silhouette width.
# Run this once on the full snapshot history to inform K selection before
# committing to a value for the backtest.
#
# Args:
#   snapshot_dat  : Monthly snapshot df from get_monthly_snapshots.
#   k_values      : Integer vector of K values to evaluate.
#   feature_cols  : Column names to use as clustering features. Defaults to
#                   momentum, vol, sharpe, and corr_spy columns (auto-detected).
#
# Returns:
#   List with $metrics (df), $p_elbow (ggplot), $p_sil (ggplot),
#   $dendrogram (hclust object for plotting).

evaluate_k <- function(snapshot_dat, k_values = 2:6, feature_cols = NULL) {

  if (is.null(feature_cols)) {
    feature_cols <- names(snapshot_dat)[
      grepl("^(momentum_|vol_[0-9]|sharpe_|corr_spy_)", names(snapshot_dat))
    ]
  }

  # Build stable sector profiles: mean feature value per sector across all dates.
  feat_mat <- snapshot_dat %>%
    dplyr::group_by(symbol) %>%
    dplyr::summarise(
      dplyr::across(dplyr::all_of(feature_cols), ~ mean(.x, na.rm = TRUE)),
      .groups = "drop"
    ) %>%
    tibble::column_to_rownames("symbol") %>%
    scale()  # normalise before distance calculation

  dist_mat   <- dist(feat_mat, method = "euclidean")
  hclust_fit <- hclust(dist_mat, method = "ward.D2")

  metrics <- purrr::map_dfr(k_values, function(k) {
    assignments <- cutree(hclust_fit, k = k)

    # Within-cluster sum of squares
    wcss <- sum(vapply(unique(assignments), function(cl) {
      members <- feat_mat[assignments == cl, , drop = FALSE]
      if (nrow(members) < 2) return(0)
      sum(apply(members, 2, var) * (nrow(members) - 1))
    }, numeric(1)))

    # Average silhouette width (requires >= 2 clusters with >= 2 members)
    sil <- tryCatch(
      mean(cluster::silhouette(assignments, dist_mat)[, "sil_width"]),
      error = function(e) NA_real_
    )

    data.frame(k = k, wcss = wcss, avg_silhouette = sil)
  })

  p_elbow <- ggplot2::ggplot(metrics, ggplot2::aes(x = k, y = wcss)) +
    ggplot2::geom_line() + ggplot2::geom_point(size = 3) +
    ggplot2::labs(title = "Elbow Method — Within-Cluster SS",
                  x = "K", y = "WCSS") +
    ggplot2::theme_minimal()

  p_sil <- ggplot2::ggplot(metrics, ggplot2::aes(x = k, y = avg_silhouette)) +
    ggplot2::geom_line() + ggplot2::geom_point(size = 3) +
    ggplot2::labs(title = "Average Silhouette Width",
                  x = "K", y = "Silhouette") +
    ggplot2::theme_minimal()

  list(metrics = metrics, p_elbow = p_elbow, p_sil = p_sil,
       dendrogram = hclust_fit)
}


# cluster_sectors: assigns cluster labels to all sectors at a single
# rebalancing date using an expanding window for feature normalisation.
#
# Expanding window: all snapshot data up to rebalance_date is used to
# compute scaling parameters (mean, sd per feature). The current snapshot's
# feature vectors are then normalised by this history before clustering.
# This ensures early rebalancing dates have less stable but still unbiased
# normalisation, which improves as the history grows.
#
# Args:
#   snapshot_dat   : Full monthly snapshot df (all dates, all symbols).
#   k              : Number of clusters to cut the dendrogram at.
#   rebalance_date : The month-end date being processed.
#   feature_cols   : Clustering feature columns. Defaults to momentum,
#                    vol, sharpe, corr_spy (auto-detected).
#
# Returns:
#   df with columns: symbol, date, cluster.

cluster_sectors <- function(snapshot_dat, k, rebalance_date,
                             feature_cols = NULL) {

  if (is.null(feature_cols)) {
    feature_cols <- names(snapshot_dat)[
      grepl("^(momentum_|vol_[0-9]|sharpe_|corr_spy_)", names(snapshot_dat))
    ]
  }

  # Expanding window: all history up to and including rebalance_date
  history <- snapshot_dat %>% dplyr::filter(date <= rebalance_date)

  # Scaling params from full history across all symbols and dates
  feat_history <- history %>%
    dplyr::select(dplyr::all_of(feature_cols)) %>%
    dplyr::filter(dplyr::if_all(dplyr::everything(), is.finite))

  feat_means <- colMeans(feat_history, na.rm = TRUE)
  feat_sds   <- apply(feat_history, 2, sd, na.rm = TRUE)

  # Current snapshot feature matrix for this rebalance date
  current <- snapshot_dat %>%
    dplyr::filter(date == rebalance_date) %>%
    dplyr::select(symbol, dplyr::all_of(feature_cols))

  if (nrow(current) == 0) return(NULL)

  feat_mat <- current %>%
    tibble::column_to_rownames("symbol") %>%
    as.matrix()

  # Apply history-derived scaling
  for (col in feature_cols) {
    if (!is.na(feat_sds[col]) && feat_sds[col] > 0) {
      feat_mat[, col] <- (feat_mat[, col] - feat_means[col]) / feat_sds[col]
    }
  }
  feat_mat[!is.finite(feat_mat)] <- 0  # replace any residual NA/Inf

  dist_mat    <- dist(feat_mat, method = "euclidean")
  hclust_fit  <- hclust(dist_mat, method = "ward.D2")
  assignments <- cutree(hclust_fit, k = k)

  data.frame(
    symbol  = names(assignments),
    date    = rebalance_date,
    cluster = as.integer(assignments),
    stringsAsFactors = FALSE
  )
}
```

```{r, "Signals", include = F}
# generate_signals: computes a continuous composite signal per sector per
# rebalancing date.
#
# Composite = MA_signal × cluster_confirmation × (1 / vol)
#
#   MA_signal            = (short_MA - long_MA) / |long_MA + epsilon|
#                          Continuous magnitude: positive = bullish trend,
#                          negative = bearish trend.
#   cluster_confirmation = fraction of cluster peers sharing the same
#                          signal direction as this sector. Amplifies
#                          sectors aligned with their cluster; penalises
#                          contrarian outliers.
#   inverse_vol          = 1 / rolling_vol.  Higher-vol sectors receive
#                          proportionally less weight.
#
# Args:
#   snapshot_dat        : Monthly snapshot df with rolling stat columns.
#   cluster_assignments : df with symbol, date, cluster (from cluster_sectors).
#   window              : Global window size used in add_rollers. Determines
#                         which columns are used as short/long MA and vol.
#
# Returns:
#   df with symbol, date, ma_signal, cluster_confirmation, composite, cluster.

generate_signals <- function(snapshot_dat, cluster_assignments, window) {

  short_w   <- max(2L, round(window / 4L))
  short_col <- paste0("mean_", short_w, "d")
  long_col  <- paste0("mean_", window, "d")
  vol_col   <- paste0("vol_", window, "d")

  snapshot_dat %>%
    dplyr::left_join(cluster_assignments, by = c("symbol", "date")) %>%
    dplyr::group_by(date) %>%
    dplyr::mutate(
      # Continuous MA crossover strength
      ma_signal = (.data[[short_col]] - .data[[long_col]]) /
                  (abs(.data[[long_col]]) + 1e-10),

      signal_dir = sign(ma_signal),

      inv_vol = dplyr::if_else(
        is.na(.data[[vol_col]]) | .data[[vol_col]] <= 0,
        NA_real_,
        1 / .data[[vol_col]]
      )
    ) %>%
    # Cluster confirmation requires grouping by cluster within each date
    dplyr::group_by(date, cluster) %>%
    dplyr::mutate(
      # For each sector: fraction of cluster members with the same direction.
      # sapply over indices so each element compares against its own direction.
      cluster_confirmation = {
        dirs <- signal_dir
        sapply(seq_along(dirs), function(i) mean(dirs == dirs[i], na.rm = TRUE))
      },
      composite = ma_signal * cluster_confirmation * inv_vol
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(symbol, date, ma_signal, cluster_confirmation,
                  composite, cluster)
}


# construct_weights: converts composite scores to dollar-neutral weights.
#
# Sectors with positive composite are held long; negative are held short.
# If all sectors share the same sign (e.g. broad bull market), a median
# split is used — top half goes long, bottom half goes short — so the
# portfolio is always dollar-neutral and always making a relative bet.
#
# Long weights sum to +1. Short weights sum to -1.
#
# Args:
#   signals_dat : df with symbol, date, composite columns.
#
# Returns:
#   df with symbol, date, weight, composite.

construct_weights <- function(signals_dat) {
  signals_dat %>%
    dplyr::group_by(date) %>%
    dplyr::mutate(
      # Median split if all composites are same sign (always stay dollar-neutral)
      med  = median(composite, na.rm = TRUE),
      is_long = dplyr::case_when(
        all(composite >= 0, na.rm = TRUE) ~ composite >= med,
        all(composite <= 0, na.rm = TRUE) ~ composite >= med,
        TRUE                              ~ composite > 0
      ),

      long_total  = sum(composite[is_long],          na.rm = TRUE),
      short_total = sum(abs(composite[!is_long]),     na.rm = TRUE),

      # Long: positive weight proportional to composite score.
      # Short: negative weight; magnitude proportional to |composite|.
      weight = dplyr::case_when(
        is_long  & long_total  > 0 ~  composite / long_total,
        !is_long & short_total > 0 ~ -abs(composite) / short_total,
        TRUE                       ~  0
      )
    ) %>%
    dplyr::ungroup() %>%
    dplyr::select(symbol, date, weight, composite)
}
```

```{r, "Backtest", include = F}
# compute_metrics: computes all performance metrics from a daily return vector.
#
# Args:
#   returns  : Numeric vector of daily strategy returns.
#   rf_daily : Daily risk-free rate. Default 0.
#
# Returns:
#   Named list: sharpe, cum_return, max_drawdown, annual_vol, omega.

compute_metrics <- function(returns, rf_daily = 0) {
  returns <- returns[is.finite(returns)]
  if (length(returns) < 20) {
    return(list(sharpe = NA_real_, cum_return = NA_real_,
                max_drawdown = NA_real_, annual_vol = NA_real_,
                omega = NA_real_, returns = returns))
  }

  excess <- returns - rf_daily

  sharpe     <- sqrt(252) * mean(excess) / sd(excess)
  cum_return <- prod(1 + returns) - 1
  annual_vol <- sd(returns) * sqrt(252)

  cumulative   <- cumprod(1 + returns)
  peak         <- cummax(cumulative)
  max_drawdown <- min((cumulative - peak) / peak)

  # Omega ratio at threshold = rf_daily
  gains  <- sum(pmax(returns - rf_daily, 0))
  losses <- sum(pmax(rf_daily - returns, 0))
  omega  <- if (losses == 0) Inf else gains / losses

  list(sharpe = sharpe, cum_return = cum_return, max_drawdown = max_drawdown,
       annual_vol = annual_vol, omega = omega, returns = returns)
}


# compute_portfolio_returns: simulates a monthly-rebalanced dollar-neutral
# portfolio with proper position drift between rebalance dates.
#
# On each execution date (the trading day after the signal date), sector
# positions are reset to target_weight * current_notional.  Between
# rebalances, positions drift freely with daily price changes — the portfolio
# is NOT re-centred to target weights until the next execution date.
#
# For a dollar-neutral portfolio, sum(weights) = 0 so the net dollar
# investment is always zero.  P&L is tracked as:
#   daily_pnl      = sum(position_i * return_i)
#   daily_return   = daily_pnl / notional
#   notional(t+1)  = notional(t) + daily_pnl
#
# where notional represents the equity capital base.
#
# Args:
#   weights_dat    : df with symbol, date, weight (output of construct_weights).
#   daily_rets_dat : df with symbol, date, ClCl.
#
# Returns:
#   df with columns date and ret (daily strategy return).

compute_portfolio_returns <- function(weights_dat, daily_rets_dat) {

  syms      <- sort(unique(weights_dat$symbol))
  all_dates <- sort(unique(daily_rets_dat$date))

  # Map each signal date to the next trading day (1-day execution lag).
  sig_dates  <- sort(unique(weights_dat$date))
  exec_dates <- vapply(sig_dates, function(d) {
    nxt <- all_dates[all_dates > d]
    if (length(nxt) == 0L) NA_real_ else as.numeric(nxt[1L])
  }, numeric(1L))
  exec_dates <- as.Date(exec_dates, origin = "1970-01-01")

  # Build a named list: "YYYY-MM-DD" -> named weight vector for fast lookup.
  exec_list <- weights_dat %>%
    dplyr::mutate(exec_date = exec_dates[match(date, sig_dates)]) %>%
    dplyr::filter(!is.na(exec_date)) %>%
    dplyr::group_by(exec_date) %>%
    dplyr::summarise(w = list(setNames(weight, symbol)), .groups = "drop")

  exec_map <- setNames(exec_list$w, as.character(exec_list$exec_date))

  # Pre-build a returns matrix (rows = dates, cols = symbols) for fast access.
  # NA returns are treated as 0 (position unchanged on that day).
  ret_mat <- daily_rets_dat %>%
    tidyr::pivot_wider(names_from = symbol, values_from = ClCl) %>%
    dplyr::arrange(date) %>%
    tibble::column_to_rownames("date") %>%
    as.matrix()
  ret_mat[is.na(ret_mat)] <- 0

  # Only begin after the first execution date.
  first_exec  <- min(exec_list$exec_date)
  trade_dates <- all_dates[all_dates >= first_exec]
  n_days      <- length(trade_dates)

  # State: dollar positions per sector and scalar equity notional.
  pos      <- setNames(numeric(length(syms)), syms)
  notional <- 1.0
  ret_vec  <- numeric(n_days)

  for (i in seq_len(n_days)) {
    d_str <- as.character(trade_dates[i])

    # On execution dates, reset positions to target weights × current notional.
    if (d_str %in% names(exec_map)) {
      new_w         <- exec_map[[d_str]]
      pos[]         <- 0
      pos[names(new_w)] <- new_w * notional
    }

    # Daily P&L = sum(position_i × return_i).
    r_today        <- ret_mat[d_str, syms]
    pnl            <- sum(pos * r_today)

    ret_vec[i]     <- pnl / notional

    # Drift positions with today's returns; accumulate P&L into notional.
    pos      <- pos * (1 + r_today)
    notional <- notional + pnl
  }

  data.frame(date = trade_dates, ret = ret_vec)
}


# run_backtest: full pipeline for a single (window, dW, k) parameter set.
#
# Pipeline:
#   add_rollers → kalman_smooth_all → monthly snapshots →
#   for each rebalance date: cluster_sectors + generate_signals +
#   construct_weights → compute_portfolio_returns → compute_metrics
#
# SPY is used for computing correlation features but excluded from the
# trading universe.
#
# Args:
#   dat         : Returns df (train_df or test_df).
#   window      : Global window size passed to add_rollers.
#   dW          : Kalman state noise passed to kalman_smooth_all.
#   k           : Number of clusters for cluster_sectors.
#   dV          : Kalman observation noise. Fixed at 1.
#   rf_daily    : Daily risk-free rate.
#   return_full : If TRUE, also returns strategy_returns_df and weights_df.
#
# Returns:
#   Output of compute_metrics plus optionally the full intermediate dfs.

run_backtest <- function(dat, window, dW, k, dV = 1, rf_daily = 0,
                         return_full = FALSE) {

  # Feature engineering — SPY stays in at this stage for corr computation
  features <- dat %>%
    add_rollers(window = window, rf_daily = rf_daily) %>%
    kalman_smooth_all(dW = dW, dV = dV)

  # Clustering features: momentum, vol, sharpe, corr_spy at the given window
  feature_cols <- c(
    paste0("momentum_", window, "d"),
    paste0("vol_",      window, "d"),
    paste0("sharpe_",   window, "d"),
    paste0("corr_spy_", window, "d")
  )

  # Monthly snapshots — exclude SPY from trading universe
  snapshots <- features %>%
    dplyr::filter(symbol != "SPY") %>%
    get_monthly_snapshots()

  # Only process rebalancing dates where all sectors have valid signals
  # (i.e., rolling window warmup is complete)
  long_col    <- paste0("mean_", window, "d")
  rebal_dates <- snapshots %>%
    dplyr::group_by(date) %>%
    dplyr::filter(all(!is.na(.data[[long_col]]))) %>%
    dplyr::pull(date) %>%
    unique() %>%
    sort()

  if (length(rebal_dates) == 0) {
    warning("No valid rebalancing dates — window may be too large for dataset.")
    return(compute_metrics(numeric(0)))
  }

  # Generate signals at each rebalancing date (expanding window clustering)
  all_signals <- purrr::map_dfr(rebal_dates, function(rd) {
    clusters <- cluster_sectors(snapshots, k = k,
                                rebalance_date = rd,
                                feature_cols = feature_cols)
    if (is.null(clusters)) return(NULL)
    generate_signals(
      snapshot_dat        = snapshots %>% dplyr::filter(date == rd),
      cluster_assignments = clusters,
      window              = window
    )
  })

  if (nrow(all_signals) == 0) return(compute_metrics(numeric(0)))

  weights <- construct_weights(all_signals)

  # Daily return series for the tradeable universe
  daily_returns <- features %>%
    dplyr::filter(symbol != "SPY") %>%
    dplyr::select(symbol, date, ClCl)

  # Simulate portfolio with proper monthly rebalancing and intra-month drift.
  strategy_returns <- compute_portfolio_returns(weights, daily_returns)

  metrics <- compute_metrics(strategy_returns$ret, rf_daily)

  if (return_full) {
    c(metrics, list(
      strategy_returns_df = strategy_returns,
      weights_df          = weights
    ))
  } else {
    metrics
  }
}
```

```{r, "Visualization", include = F}
# plot_backtest: performance tearsheet for a single backtest run.
# Requires the $returns vector in the backtest result list.
#
# Args:
#   backtest_result : Output from run_backtest (must include $returns).
#   title           : Chart title string.
#
# Returns:
#   A patchwork-combined ggplot of 4 panels: cumulative return, drawdown,
#   rolling 60-day Sharpe, rolling 60-day annualised volatility.

plot_backtest <- function(backtest_result,
                          title = "Strategy Performance") {

  returns <- backtest_result$returns
  if (length(returns) == 0 || all(is.na(returns))) {
    stop("backtest_result$returns is empty. Run run_backtest() first.")
  }

  n  <- length(returns)
  cr <- cumprod(1 + returns)

  df <- data.frame(
    t        = seq_len(n),
    ret      = returns,
    cum_ret  = cr - 1,
    drawdown = (cr - cummax(cr)) / cummax(cr),
    roll_sharpe = slider::slide_dbl(
      returns, ~ sqrt(252) * mean(.x) / sd(.x),
      .before = 59, .complete = TRUE
    ),
    roll_vol = slider::slide_dbl(
      returns, ~ sd(.x) * sqrt(252),
      .before = 59, .complete = TRUE
    )
  )

  p1 <- ggplot2::ggplot(df, ggplot2::aes(x = t, y = cum_ret)) +
    ggplot2::geom_line(color = "#2196F3", linewidth = 0.7) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
    ggplot2::labs(title = "Cumulative Return", x = NULL, y = NULL) +
    ggplot2::theme_minimal()

  p2 <- ggplot2::ggplot(df, ggplot2::aes(x = t, y = drawdown)) +
    ggplot2::geom_area(fill = "#F44336", alpha = 0.55) +
    ggplot2::labs(title = "Drawdown", x = NULL, y = NULL) +
    ggplot2::theme_minimal()

  p3 <- ggplot2::ggplot(df, ggplot2::aes(x = t, y = roll_sharpe)) +
    ggplot2::geom_line(color = "#4CAF50", linewidth = 0.7) +
    ggplot2::geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.4) +
    ggplot2::labs(title = "Rolling 60-Day Sharpe", x = NULL, y = NULL) +
    ggplot2::theme_minimal()

  p4 <- ggplot2::ggplot(df, ggplot2::aes(x = t, y = roll_vol)) +
    ggplot2::geom_line(color = "#FF9800", linewidth = 0.7) +
    ggplot2::labs(title = "Rolling 60-Day Vol (Ann.)",
                  x = "Trading Day", y = NULL) +
    ggplot2::theme_minimal()

  (p1 / p2 / p3 / p4) +
    patchwork::plot_annotation(title = title)
}
```

# Introduction

This document outlines the design, implementation, and evaluation of a momentum-based sector rotation strategy applied to the eleven SPDR S&P 500 sector ETFs alongside SPY as a benchmark. The strategy uses rolling feature engineering, Kalman filter smoothing, hierarchical clustering, and a composite MA-crossover signal to construct a monthly-rebalanced, dollar-neutral long-short portfolio.

The pipeline consists of four main stages: (1) computation of rolling statistics for each sector, (2) Kalman smoothing of those features to reduce noise, (3) hierarchical clustering of sectors at each monthly rebalance date, and (4) construction of a composite signal that drives weight allocation. A one-day execution lag is applied throughout to avoid same-bar execution bias.

The training period covers mid-2000 through December 2024. Testing runs from January 2025 through February 2026. Baseline parameters (window = 126, dW = 0.01, K = 3) were selected heuristically and later evaluated against a broader parameter grid.

------------------------------------------------------------------------

## Trading Universe

The strategy trades eleven SPDR S&P 500 sector ETFs. SPY is included in data pulls to compute correlation features but is excluded from the trading universe itself.

```{r, "Universe Table Compute", include = F}
universe_info <- tibble::tribble(
  ~Ticker, ~Sector,
  "XLB",  "Materials",
  "XLC",  "Communication Services",
  "XLE",  "Energy",
  "XLF",  "Financials",
  "XLI",  "Industrials",
  "XLK",  "Information Technology",
  "XLP",  "Consumer Staples",
  "XLRE", "Real Estate",
  "XLV",  "Health Care",
  "XLU",  "Utilities",
  "XLY",  "Consumer Discretionary"
)

data_avail <- raw_data %>%
  dplyr::filter(symbol != "SPY") %>%
  dplyr::group_by(symbol) %>%
  dplyr::summarise(
    `Data From` = format(min(date), "%Y-%m-%d"),
    `Data To`   = format(max(date), "%Y-%m-%d"),
    .groups = "drop"
  ) %>%
  dplyr::rename(Ticker = symbol)

universe_tbl <- universe_info %>%
  dplyr::left_join(data_avail, by = "Ticker")
```

```{r, "Universe Table - Out", echo = F, warning = F, message = F}
universe_tbl %>%
  gt::gt() %>%
  gt::tab_header(
    title    = "Trading Universe",
    subtitle = "SPDR S&P 500 Sector ETFs"
  ) %>%
  gt::cols_label(
    Ticker    = gt::md("**Ticker**"),
    Sector    = gt::md("**Sector**"),
    `Data From` = gt::md("**Data From**"),
    `Data To`   = gt::md("**Data To**")
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f5f5f5"),
    locations = gt::cells_body(rows = seq(1, nrow(universe_tbl), by = 2))
  ) %>%
  gt::tab_footnote(
    footnote  = "XLC launched June 2018; XLRE launched October 2015. All other ETFs have data from the pull start date of 2000-06-01.",
    locations = gt::cells_column_labels(columns = `Data From`)
  ) %>%
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) %>%
  gt::tab_options(table.width = gt::pct(75))
```

------------------------------------------------------------------------

# Rationale

The academic support for momentum as a persistent return predictor is substantial. Jegadeesh and Titman (1993) were among the first to rigorously document return continuation across U.S. equities, finding that portfolios of prior winners consistently outperformed prior losers over 3–12 month horizons. This anomaly has proven durable: Asness, Moskowitz, and Pedersen (2013) documented its persistence across multiple asset classes and geographies over long out-of-sample periods.

More directly relevant to this strategy, Moskowitz and Grinblatt (1999) showed that industry-level momentum accounts for a significant portion of the individual stock momentum effect. Operating at the sector ETF level rather than on individual securities attempts to capture exactly this industry component while benefiting from the liquidity and diversification that broad sector baskets provide.

A known risk in any momentum approach is the momentum crash phenomenon described by Daniel and Moskowitz (2016), where momentum strategies experience sharp reversals, typically after periods of high market stress and subsequent recovery. The strategy described here has no explicit crash protection, which is a meaningful unaddressed risk.

## Correlation

To justify a cross-sector relative-value strategy, the return correlation structure across sectors must be examined. High correlation confirms common market-wide exposure that a dollar-neutral position will partially hedge, while meaningful pairwise variation across sectors provides the information content needed for relative bets.

```{r, "Correlation Compute", include = F}
cor_mat <- train_df %>%
  dplyr::select(symbol, date, ClCl) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = ClCl) %>%
  dplyr::select(-date) %>%
  cor(use = "complete.obs")

cor_long <- as.data.frame(as.table(cor_mat)) %>%
  dplyr::rename(Asset1 = Var1, Asset2 = Var2, Correlation = Freq)

p_corr <- ggplot2::ggplot(cor_long, ggplot2::aes(x = Asset1, y = Asset2, fill = Correlation)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::geom_text(ggplot2::aes(label = round(Correlation, 2)), size = 2.8) +
  ggplot2::scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                                 midpoint = 0, limits = c(-1, 1)) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(title = "Sector ETF Return Correlation", x = NULL, y = NULL)
```

```{r, "Correlation - Out", echo = F, warning = F, message = F}
p_corr
```

The correlation matrix confirms substantial co-movement across all sector ETFs, as expected given shared exposure to the broad equity market. Correlations are uniformly positive, with most pairs in the 0.6–0.9 range. Defensive sectors (XLU, XLRE, XLP) tend to correlate slightly less with cyclicals (XLK, XLY, XLE), which reflects their different factor sensitivities and provides some basis for relative positioning. A dollar-neutral portfolio will hedge out a large fraction of market-wide risk, but correlated drawdowns during severe market stress will still affect both long and short legs.

------------------------------------------------------------------------

# Assumptions

The backtest results throughout this document rest on several simplifying assumptions that would not hold in live trading. Each omission tends to flatter the reported returns.

**No transaction costs.** No commissions, exchange fees, or clearing costs are deducted from returns. Even small per-trade costs compound significantly over a 24-year backtest. At a conservative estimate of \$5 per sector trade, monthly rebalancing across 11 sectors would incur over \$16,000 in costs per year on a \$1M book, before accounting for the short leg.

**No bid-ask spread.** All trades are assumed to execute at the prevailing closing price rather than the ask (for buys) or bid (for sells). The SPDR sector ETFs are highly liquid with spreads of 1–2 cents, so the per-trade impact is small, but it is non-zero and systematically omitted.

**No short-selling costs.** Short positions require borrowing shares, which incurs a daily fee charged by the prime broker. For the SPDR ETFs this is typically under 0.1% per year, but the cost is not constant and can spike when short interest concentrates in a particular sector.

**No market impact.** Orders are assumed not to move the market price. This is reasonable for a small portfolio trading highly liquid ETFs, but would break down at institutional scale where a large sector-ETF order can meaningfully push prices.

**No ETF expense ratios.** Each SPDR sector ETF carries a gross expense ratio (approximately 0.09–0.13% per year). These fees are embedded in the adjusted price series and are therefore captured in the return calculations, but any fund-level management fee sitting on top of the underlying ETFs would not be.

**Exact 1-day execution lag.** The strategy uses a 1-day lag to simulate executing signals the day after they are observed. In practice, execution may span multiple days depending on order size, market conditions, and broker constraints.

**No taxes.** Capital gains taxes, wash-sale rules, and the distinction between short-term and long-term treatment are not modelled. Monthly rebalancing creates frequent realisation events, which would typically produce short-term capital gains — taxed at ordinary income rates — representing a substantial drag in a taxable account.

**Full dividend reinvestment at zero cost.** Returns are computed on adjusted close prices, which assumes dividends are reinvested on the ex-dividend date instantaneously and without cost or tax. This is standard in research but slightly overstates real-world performance.

**No margin or leverage cost.** The dollar-neutral construction uses 2:1 gross leverage (long \$1 and short \$1 per \$1 of capital). No cost of borrowing the margin, and no margin call risk, is modelled.

**Unlimited short availability.** The strategy assumes shares can always be borrowed to establish short positions. While SPDR sector ETFs rarely experience short squeezes or hard-to-borrow conditions, this assumption is not guaranteed.

------------------------------------------------------------------------

# Strategy

## Overview

```{mermaid}
flowchart TD
    A[Yahoo Price Data] --> B[Pull & Align Returns Sector ETFs + SPY]
    B --> C[Feature Engineering MA · Vol · Sharpe · Momentum · Corr · Drawdown]
    C --> D[Kalman Smoothing Normalise → Local-Level Filter controlled by dW]
    D --> E[Monthly Snapshots Expanding-window normalise at each rebalance date]
    E --> F[Hierarchical Clustering Ward linkage · K clusters Peer-group assignment]
    F --> G[Signal Construction MA crossover × cluster confirmation × inverse volatility]
    G --> H[Weight Allocation Long / Short · Dollar-neutral 1-day execution lag]
    H --> I[Daily P&L ClCl returns × lagged weights]
    I --> J[Performance Metrics Sharpe · Cum Return · Drawdown Vol · Omega]

    subgraph Optimisation
        K[expand.grid window × dW] --> L[foreach %dopar% run_backtest per combo]
        L --> J
    end

    style A fill:#4a4a6a,color:#fff
    style J fill:#2d6a4f,color:#fff
    style Optimisation fill:#1a1a2e,color:#fff
```

The strategy pipeline has four stages:

**1. Feature Engineering:** Rolling statistics are computed daily per sector: short and long moving averages (for the MA crossover signal), rolling volatility (standard deviation of ClCl returns), Sharpe ratio, momentum (cumulative compound return), correlation to SPY, and maximum drawdown. The baseline uses a 126-day global window with a short window of 32 days (approximately window/4).

**2. Kalman Smoothing:** Each rolling feature series is passed through a local-level Kalman filter (random-walk state, Gaussian observation noise). Series are normalised to zero mean and unit variance before filtering so that the single state-noise parameter *dW* is comparable across all features. Lower *dW* produces heavier smoothing and slower adaptation to regime changes; higher *dW* tracks the raw data more closely.

**3. Hierarchical Clustering:** At each monthly rebalancing date, the eleven sector ETFs are grouped into K clusters using hierarchical clustering (Ward's linkage, Euclidean distance) applied to the smoothed feature vectors. An expanding-window normalisation is used at each rebalance date to avoid look-ahead bias. The intent is to identify peer groups (sectors behaving similarly) whose consensus can confirm or penalise individual signals.

**4. Signal Construction and Weight Allocation:** A composite signal is computed per sector as the product of: (a) the continuous MA crossover magnitude `(short_MA - long_MA) / |long_MA|`, (b) a cluster confirmation score (fraction of cluster peers sharing the same directional signal), and (c) inverse volatility. Sectors with a positive composite are held long; negative composites are held short. Weights are proportional to composite magnitude within each side. The portfolio is dollar-neutral at every rebalance. A one-day execution lag is applied: signals observed at month-end close are acted on the following trading day.

------------------------------------------------------------------------

## Kalman Filtering

The Kalman filter used here is a local-level model, sometimes called a random-walk-plus-noise model. It treats each feature series (rolling momentum, rolling Sharpe, etc.) as a noisy observation of a true underlying state, where that state evolves slowly over time as a random walk.

The model has two equations. The state equation says that the true value at time t equals the true value at time t-1 plus a small random disturbance with variance dW. The observation equation says that what we observe equals the true state plus independent noise with variance dV. Written out:

-   State: μ_t = μ\_{t-1} + w_t, where w_t \~ N(0, dW)
-   Observation: y_t = μ_t + v_t, where v_t \~ N(0, dV)

At each time step, the filter produces an estimate of the true state by combining two sources of information: the prediction from the previous state, and the new observation. The relative weight given to each depends on the ratio dW/dV. When dW is small relative to dV, the filter treats observations as noisy and trusts the prediction more, producing a heavily smoothed output that changes slowly. When dW is large relative to dV, the filter trusts each new observation more and updates quickly, tracking the raw series closely.

In this implementation, dV is fixed at 1. Before filtering, each series is normalised to zero mean and unit variance, so the same dW value applies meaningfully across all features regardless of their original scale. The baseline dW of 0.01 means the state is assumed to evolve slowly relative to observation noise, which produces substantial smoothing across all features.

The filter runs strictly forward: each output value depends only on past observations. This makes it causal and prevents look-ahead bias. The trade-off is that an exponential moving average (EMA) applied to the rolling statistics would achieve comparable smoothing with a single interpretable decay parameter and no distributional assumptions. The Kalman approach is theoretically cleaner but adds complexity without a demonstrated empirical advantage in this specific context.

------------------------------------------------------------------------

## Shorting and Dollar-Neutrality

A short position profits when an asset's price falls. The strategy borrows shares of a sector ETF from a broker, sells them immediately at the current market price, and later repurchases them at a (hoped-for) lower price. The profit is the difference between the initial sale and the repurchase, minus the cost of borrowing. In the context of this strategy, a sector receives a negative weight when its composite signal is negative, indicating it is expected to underperform peers. Multiplying a negative weight by a negative return (the sector fell) produces a positive contribution to portfolio return.

Dollar-neutrality means the total dollar value of long positions equals the total dollar value of short positions, resulting in zero net market exposure. In this implementation, long weights sum to +1 and short weights sum to -1 at every rebalancing date. The portfolio is simultaneously \$1 long a basket of sector ETFs and \$1 short another basket, for a net invested position of zero.

The primary benefit is market-beta removal. Because both the long and short legs hold sector ETFs with high pairwise correlation, broad market moves tend to affect both legs in the same direction and roughly the same magnitude, largely cancelling out. The portfolio profits only when the long sectors outperform the short sectors on a relative basis, regardless of whether the overall market goes up or down. This is why the strategy is described as a relative-value long-short strategy rather than a directional one.

Dollar-neutrality does not eliminate all risk. The correlation matrix shows that all sector pairs are positively correlated, which means both legs can lose simultaneously in severe market dislocations. The hedge is imperfect because sector correlations are not exactly equal, and factor exposures (e.g., to value, growth, or interest-rate sensitivity) can differ substantially across sectors.

When all composite scores carry the same sign (for example, all sectors showing a positive trend signal in a broad bull market), a strict positive/negative split would leave one side empty. To handle this, the strategy applies a median split: sectors above the median composite score go long, sectors below go short. This ensures the portfolio is always dollar-neutral and always making a relative bet between stronger and weaker performers, even when the signal has a consistent directional bias.

------------------------------------------------------------------------

## Signal Logic

The composite signal has three multiplicative components:

**composite = MA_signal × cluster_confirmation × (1 / vol)**

**MA Signal** is the normalised distance between the short and long moving averages. A positive value means the short MA is above the long MA (uptrend); a negative value means the short MA is below the long MA (downtrend). The denominator normalises by the magnitude of the long MA so that the signal is comparable across sectors and time periods regardless of absolute price level.

**Cluster Confirmation** is the fraction of cluster peers sharing the same signal direction. A sector whose trend aligns with its peers scores close to 1.0; a contrarian sector that is the only one in its cluster pointing in a given direction scores close to 0.0. This multiplicative penalty reduces the weight of outlier signals and amplifies signals that are consistent across the peer group.

**Inverse Volatility** scales positions down in high-volatility sectors and up in low-volatility sectors. All else equal, a sector with half the volatility receives twice the weight, which is standard in risk-parity-style weighting.

The table below summarises how combinations of these components map to trade decisions and relative position sizes.

| MA Signal | Cluster Confirmation | Volatility | Composite | Position |
|---------------|---------------|---------------|---------------|---------------|
| Short MA \> long MA (uptrend) | High (peers agree) | Low | Large positive | **Long, high weight** |
| Short MA \> long MA (uptrend) | Low (peers disagree) | Any | Small positive | Long, reduced weight |
| Short MA \< long MA (downtrend) | High (peers agree) | Low | Large negative | **Short, high weight** |
| Short MA \< long MA (downtrend) | Low (peers disagree) | Any | Small negative | Short, reduced weight |
| All sectors same direction | N/A | Any | Median split applied | Top half long, bottom half short |

A concrete example: XLE has a short MA above its long MA (positive trend), its cluster peers XLB and XLI share the same direction (confirmation ≈ 1.0), and it has below-average volatility. It receives a large positive composite and is assigned a high long weight. Meanwhile, XLC has a short MA below its long MA, its peers agree, but it has above-average volatility. It receives a moderate negative composite and is assigned a modest short weight.

------------------------------------------------------------------------

## Clustering

To select K, the dendrogram and elbow/silhouette diagnostics are examined over K = 2 through 6. Sectors are characterised by their mean feature values across all training snapshots and clustered once to produce stable groupings for visual inspection.

```{r, "K Evaluation Compute", include = F}
baseline_window <- 126

baseline_features <- train_df %>%
  add_rollers(window = baseline_window) %>%
  kalman_smooth_all(dW = 0.01)

train_snapshots <- baseline_features %>%
  dplyr::filter(symbol != "SPY") %>%
  get_monthly_snapshots()

k_eval <- evaluate_k(train_snapshots, k_values = 2:6)
```

```{r, "K Evaluation - Out", echo = F, warning = F, message = F}
k_eval$p_elbow + k_eval$p_sil
```

```{r, "Dendrogram - Out", echo = F, warning = F, message = F}
plot(k_eval$dendrogram, main = "Sector Dendrogram", sub = "", xlab = "")
```

The diagnostics suggest K = 3 as a reasonable baseline. The silhouette peaks there and the elbow in the WCSS curve is most visible at that point, though the differences across K = 2 to 4 are modest given the small universe of eleven sectors. The dendrogram broadly separates defensive/income sectors (XLU, XLRE, XLP), growth/technology sectors (XLK, XLC, XLY), and cyclical/value sectors (XLE, XLB, XLI, XLF, XLV). This grouping is broadly consistent with how practitioners classify sectors, which provides a basic sanity check that the clustering is picking up real structure rather than noise.

It is worth noting that with only eleven sectors, any clustering algorithm will produce coarse groupings. The marginal value of the clustering step over a manually-defined sector-group categorisation has not been rigorously tested here.

------------------------------------------------------------------------

# Performance

The strategy was evaluated on two periods: a long in-sample training period (June 2000 – December 2024) and a short out-of-sample testing period (January 2025 – February 2026). Baseline parameters of window = 126, dW = 0.01, and K = 3 were used for both runs. These parameters were selected after observing training-period behaviour, so in-sample metrics should be interpreted cautiously.

------------------------------------------------------------------------

## Training

The training run covers approximately 24 years of sector ETF history. This period includes the dot-com bust, the 2008 financial crisis, the 2020 COVID selloff, and the subsequent recovery, covering a wide range of market conditions. A long history provides statistical stability, but it also means the strategy will have experienced multiple momentum regime shifts, including the crashes documented by Daniel and Moskowitz (2016).

```{r, "Training Backtest Compute", include = F}
baseline_train <- run_backtest(train_df, window = 126, dW = 0.01, k = 3,
                               return_full = TRUE)
```

```{r, "XTS Prep - Training", include = F}
# Core daily equity series
tick_train <- baseline_train$strategy_returns_df %>%
  dplyr::mutate(
    cumeq       = cumprod(1 + ret),
    drawdown    = (cumeq - cummax(cumeq)) / cummax(cumeq),
    roll_sharpe = slider::slide_dbl(ret, ~ sqrt(252) * mean(.x) / sd(.x),
                                    .before = 59, .complete = TRUE)
  ) %>%
  timetk::tk_xts(date_var = date)

# SPY benchmark cumulative return over same period
spy_bench_train <- train_df %>%
  dplyr::filter(
    symbol == "SPY",
    date   >= min(baseline_train$strategy_returns_df$date),
    date   <= max(baseline_train$strategy_returns_df$date)
  ) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(spy_cumeq = cumprod(1 + ClCl)) %>%
  dplyr::select(date, spy_cumeq) %>%
  timetk::tk_xts(date_var = date)

# Top and bottom weight per rebalance date, forward-filled to daily
top_bot_train <- baseline_train$weights_df %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(top_weight = max(weight, na.rm = TRUE),
                   bot_weight = min(weight, na.rm = TRUE),
                   .groups = "drop")

tick_exposure_train <- baseline_train$strategy_returns_df %>%
  dplyr::select(date) %>%
  dplyr::left_join(top_bot_train, by = "date") %>%
  dplyr::arrange(date) %>%
  tidyr::fill(top_weight, bot_weight, .direction = "down") %>%
  timetk::tk_xts(date_var = date)

# Per-sector weights at each monthly rebalance as XTS (for multi-line plot)
tick_sector_weights_train <- baseline_train$weights_df %>%
  dplyr::select(symbol, date, weight) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = weight) %>%
  timetk::tk_xts(date_var = date)

# Sector weight heatmap (rebalance dates only)
p_sector_heatmap_train <- baseline_train$weights_df %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = symbol, fill = weight)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::scale_fill_gradient2(low = "#F44336", mid = "white", high = "#2196F3",
                                 midpoint = 0) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(title = "Sector Weights at Each Rebalance — Training",
                x = NULL, y = NULL)
```

```{r, "Training - Tearsheet", echo = F, warning = F, message = F}
plot_backtest(baseline_train, title = "Training Period — Strategy Performance")
```

```{r, "Training Chart B - Strategy vs SPY", echo = F, warning = F, message = F, include = F}
plot(tick_train$cumeq, main = "Training — Strategy vs SPY")
```

```{r, "Training Chart B - SPY CumEq", echo = F, warning = F, message = F, include = F}
xts::addSeries(spy_bench_train$spy_cumeq,
               main = "SPY CumEq", on = NA,
               type = "l", col = "orange", lty = 1, lwd = 1, pch = 0)
```

```{r, "Training Chart B - Drawdown", echo = F, warning = F, message = F}
xts::addSeries(tick_train$drawdown,
               main = "Drawdown", on = NA,
               type = "l", col = "red", lty = 1, lwd = 1, pch = 0)
```

```{r, "Training Chart C - CumEQ Equity", echo = F, warning = F, message = F, include = F}
plot(tick_train$cumeq, main = "Training — Equity & Rolling Sharpe")
```

```{r, "Training Chart C - Rolling Sharpe", echo = F, warning = F, message = F, include = F}
xts::addSeries(tick_train$roll_sharpe,
               main = "Rolling 60d Sharpe", on = NA,
               type = "l", col = "purple", lty = 1, lwd = 1, pch = 0)
```

```{r, "Training Chart C - Daily Returns", echo = F, warning = F, message = F}
xts::addSeries(tick_train$ret,
               main = "Daily Returns", on = NA,
               type = "h", col = "blue", lty = 1, lwd = 1, pch = 0)
```

```{r, "Training Chart E - Sector Weight Heatmap", echo = F, warning = F, message = F}
p_sector_heatmap_train
```

The training period results are negative: the strategy lost money over the 2000–2024 period. The cumulative equity curve ends below its starting value, and the Sharpe ratio is below zero. This is a poor result, and it is the in-sample result, meaning there was no out-of-sample degradation needed to produce it; the strategy simply did not work on the data it was designed on.

The equity curve does show periods of gains, but these were offset by drawdowns that were not recovered. The strategy appears to respond to some momentum regimes but reverses sharply in others, with no net positive edge over the full period.

The SPY comparison makes the result starker. A long-only SPY position over the same period, despite living through two major crashes, significantly outperformed the strategy. The dollar-neutral design is intended to hedge market exposure, but it did not convert that hedge into a positive return; it simply removed market beta while the relative sector bets also failed to add value on net.

------------------------------------------------------------------------

## Testing

The testing period covers January 2025 through February 2026, approximately 14 months. This is a fairly short window, and any conclusions drawn from it should be held loosely. One poor or good run in a short period can look dramatic on a cumulative return chart without being statistically meaningful.

```{r, "Testing Backtest Compute", include = F}
baseline_test <- run_backtest(test_df, window = 126, dW = 0.01, k = 3,
                              return_full = TRUE)
```

```{r, "XTS Prep - Testing", include = F}
tick_test <- baseline_test$strategy_returns_df %>%
  dplyr::mutate(
    cumeq       = cumprod(1 + ret),
    drawdown    = (cumeq - cummax(cumeq)) / cummax(cumeq),
    roll_sharpe = slider::slide_dbl(ret, ~ sqrt(252) * mean(.x) / sd(.x),
                                    .before = 59, .complete = TRUE)
  ) %>%
  timetk::tk_xts(date_var = date)

spy_bench_test <- test_df %>%
  dplyr::filter(
    symbol == "SPY",
    date   >= min(baseline_test$strategy_returns_df$date),
    date   <= max(baseline_test$strategy_returns_df$date)
  ) %>%
  dplyr::arrange(date) %>%
  dplyr::mutate(spy_cumeq = cumprod(1 + ClCl)) %>%
  dplyr::select(date, spy_cumeq) %>%
  timetk::tk_xts(date_var = date)

top_bot_test <- baseline_test$weights_df %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(top_weight = max(weight, na.rm = TRUE),
                   bot_weight = min(weight, na.rm = TRUE),
                   .groups = "drop")

tick_exposure_test <- baseline_test$strategy_returns_df %>%
  dplyr::select(date) %>%
  dplyr::left_join(top_bot_test, by = "date") %>%
  dplyr::arrange(date) %>%
  tidyr::fill(top_weight, bot_weight, .direction = "down") %>%
  timetk::tk_xts(date_var = date)

tick_sector_weights_test <- baseline_test$weights_df %>%
  dplyr::select(symbol, date, weight) %>%
  tidyr::pivot_wider(names_from = symbol, values_from = weight) %>%
  timetk::tk_xts(date_var = date)

p_sector_heatmap_test <- baseline_test$weights_df %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = symbol, fill = weight)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::scale_fill_gradient2(low = "#F44336", mid = "white", high = "#2196F3",
                                 midpoint = 0) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  ggplot2::labs(title = "Sector Weights at Each Rebalance — Testing",
                x = NULL, y = NULL)
```

```{r, "Testing - Tearsheet", echo = F, warning = F, message = F}
plot_backtest(baseline_test, title = "Testing Period — Strategy Performance")
```

```{r, "Testing Chart A - CumEQ", echo = F, warning = F, message = F, include = F}
plot(tick_test$cumeq, main = "Testing — Cumulative Equity")
```

```{r, "Testing Chart B - Strategy vs SPY", echo = F, warning = F, message = F, include = F}
plot(tick_test$cumeq, main = "Testing — Strategy vs SPY")
```

```{r, "Testing Chart B - SPY CumEq", echo = F, warning = F, message = F, include = F}
xts::addSeries(spy_bench_test$spy_cumeq,
               main = "SPY CumEq", on = NA,
               type = "l", col = "orange", lty = 1, lwd = 1, pch = 0)
```

```{r, "Testing Chart B - Drawdown", echo = F, warning = F, message = F}
xts::addSeries(tick_test$drawdown,
               main = "Drawdown", on = NA,
               type = "l", col = "red", lty = 1, lwd = 1, pch = 0)
```

```{r, "Testing Chart D - Sector Weight Heatmap", echo = F, warning = F, message = F}
p_sector_heatmap_test
```

The testing period performance is also negative, which is consistent with but no worse than the training result. Since the strategy lost money in-sample, continued losses out-of-sample are not a surprise. The signal simply does not appear to have a positive edge in the configuration tested.

The sector weight heatmap shows the model continued to rotate positions, which confirms the mechanics are running correctly. The issue is not a broken pipeline; weights are being assigned and sectors are being traded, but the direction of those trades was not profitable on net.

The 14-month testing window is too short to draw statistical conclusions in isolation. But combined with the negative training result, there is no evidence from either period that the current signal construction generates a reliable edge.

------------------------------------------------------------------------

## Strategy vs SPY

```{r, "Comparison Compute", include = F}
# Build a long-format dataframe of cumulative equity for strategy and SPY
# across both periods. Both series are indexed to 1 at the first strategy date.

make_comparison_df <- function(backtest_result, price_df, period_label) {
  strat_dates <- backtest_result$strategy_returns_df$date

  strat <- backtest_result$strategy_returns_df %>%
    dplyr::mutate(
      cumeq  = cumprod(1 + ret),
      series = "Strategy",
      period = period_label
    ) %>%
    dplyr::select(date, value = cumeq, series, period)

  spy <- price_df %>%
    dplyr::filter(
      symbol == "SPY",
      date   >= min(strat_dates),
      date   <= max(strat_dates)
    ) %>%
    dplyr::arrange(date) %>%
    dplyr::mutate(
      cumeq  = cumprod(1 + ClCl),
      series = "SPY",
      period = period_label
    ) %>%
    dplyr::select(date, value = cumeq, series, period)

  dplyr::bind_rows(strat, spy)
}

compare_df <- dplyr::bind_rows(
  make_comparison_df(baseline_train, train_df, "Training"),
  make_comparison_df(baseline_test,  test_df,  "Testing")
) %>%
  dplyr::mutate(period = factor(period, levels = c("Training", "Testing")))

p_compare_train <- compare_df %>%
  dplyr::filter(period == "Training") %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = value, color = series)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.35) +
  ggplot2::scale_color_manual(values = c("Strategy" = "#2196F3", "SPY" = "#FF9800")) +
  ggplot2::scale_y_continuous(labels = scales::label_number(suffix = "x")) +
  ggplot2::labs(title = "Training Period (2000–2024)", x = NULL,
                y = "Cumulative Return (indexed to 1)", color = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")

p_compare_test <- compare_df %>%
  dplyr::filter(period == "Testing") %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = value, color = series)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed", alpha = 0.35) +
  ggplot2::scale_color_manual(values = c("Strategy" = "#2196F3", "SPY" = "#FF9800")) +
  ggplot2::scale_y_continuous(labels = scales::label_number(suffix = "x")) +
  ggplot2::labs(title = "Testing Period (2025–2026)", x = NULL,
                y = "Cumulative Return (indexed to 1)", color = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

```{r, "Comparison - Out", echo = F, warning = F, message = F}
(p_compare_train / p_compare_test) +
  patchwork::plot_annotation(
    title   = "Strategy vs SPY: Cumulative Return",
    caption = "Both series indexed to 1 at the first strategy rebalancing date. Blue = strategy, orange = SPY buy-and-hold."
  )
```

The two panels show clearly how the strategy performed relative to a passive SPY position in each period. During training, SPY compounded substantially over the 24-year window, while the strategy's cumulative return drifted negative. During testing, both the strategy and SPY faced the 2025 drawdown period, but the SPY comparison makes plain that the dollar-neutral design did not provide the relative return needed to justify its complexity against a simple long-only benchmark.

------------------------------------------------------------------------

# Optimization

To assess parameter sensitivity and move beyond heuristically-chosen baseline values, a continuous grid search was run over the following ranges with K fixed at 3:

-   **Window:** 20 to 252 days (step 5), covering the full range from one month to one year of trading days
-   **dW (Kalman state noise):** 0.001 to 1.0 (step 0.05)

This produced \~940 combinations, evaluated on the training period only. The goal is sensitivity analysis: identifying whether there is a robust region of the parameter space that performs consistently, or whether results are noise-driven.

```{r, "Optimization Compute", include = F}
vars_to_test <- expand.grid(
  window = seq(20, 252, by = 10),
  dW     = seq(0.001, 1, by = 0.05),
  stringsAsFactors = FALSE
)

n_cores <- detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

opt_results <- foreach(
  i         = 1:nrow(vars_to_test),
  .combine  = "cbind",
  .packages = c("dplyr", "dlm", "slider", "tidyr", "purrr", "cluster", "tibble")
) %dopar% {
  p      <- vars_to_test[i, ]
  result <- tryCatch(
    run_backtest(train_df, window = p$window, dW = p$dW, k = 3),
    error = function(e) list(sharpe = NA_real_, cum_return = NA_real_,
                              max_drawdown = NA_real_, annual_vol = NA_real_,
                              omega = NA_real_)
  )
  as.numeric(c(result$sharpe, result$cum_return, result$max_drawdown,
               result$annual_vol, result$omega))
}

stopCluster(cl)

opt_results           <- tibble::as_tibble(t(opt_results))
colnames(opt_results) <- c("sharpe", "cum_return", "max_drawdown", "annual_vol", "omega")
opt_results           <- cbind(vars_to_test, opt_results)
```

```{r, "Optimization 3D Surface", include = F}
opt_window_vals <- unique(opt_results$window)
opt_dW_vals     <- unique(opt_results$dW)

opt_results <- opt_results %>%
  dplyr::mutate(log_cum_return = log(pmax(cum_return + 1, 1e-6)))

cum_ret_matrix <- opt_results %>%
  dplyr::select(window, dW, log_cum_return) %>%
  tidyr::pivot_wider(values_from = log_cum_return, names_from = dW) %>%
  dplyr::arrange(window) %>%
  dplyr::select(-window) %>%
  as.matrix()

opt_3d_plot <- plot_ly(x = ~opt_window_vals, y = ~opt_dW_vals, z = ~cum_ret_matrix) %>%
  add_surface() %>%
  layout(
    scene = list(
      xaxis = list(title = "Window"),
      yaxis = list(title = "dW"),
      zaxis = list(title = "Log Cum. Return")
    ),
    title = "Optimization Surface — Log Cumulative Return"
  )
```

```{r, "Optimization Heatmap", include = F}
opt_long <- opt_results %>%
  stats::na.omit() %>%
  tidyr::pivot_longer(cols = -c(window, dW), names_to = "variable", values_to = "value") %>%
  dplyr::group_by(variable) %>%
  dplyr::mutate(valueZ = (value - mean(value)) / sd(value))

opt_heatmap <- ggplot(opt_long, aes(x = as.numeric(window), y = as.numeric(dW))) +
  geom_raster(aes(fill = valueZ), interpolate = TRUE) +
  facet_wrap(~ variable, scales = "free") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue",
                       midpoint = median(opt_long$valueZ)) +
  scale_x_continuous(n.breaks = 5) +
  scale_y_continuous(n.breaks = 5) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.background = element_rect(fill = "white")
  ) +
  labs(title = "Optimization Results Grid", subtitle = "Z-score for Comparability",
       x = "Window", y = "dW")
```

```{r, "Optimization - Out", echo = F, warning = F, message = F}
opt_3d_plot
opt_heatmap
```

The optimization results show whether losses are broadly consistent across parameter combinations or whether a robust region exists. If no coherent optimum appears in the surface, it suggests the returns are noise-driven rather than reflecting a stable underlying signal that happens to be misconfigured.

------------------------------------------------------------------------

# Lessons Learned

## Data and Feature Engineering

Pulling historical sector ETF data via `tidyquant` was operationally straightforward, but the rolling statistics function grew in complexity as features were added. Keeping track of dynamic column naming (e.g., `mean_32d`, `vol_126d`) across a generic window parameter required careful convention management, and mismatches between expected and actual column names caused silent bugs early in development.

An error made in an early version of the returns code was using raw `close` prices for overnight return calculations rather than `adjusted` values. For a same-day `OpCl` return this is correct, but for `ClCl` (total daily return) and `ClOp` (overnight), `adjusted` close must be used to account for splits and dividends. This caused performance figures to look subtly wrong until caught. The fix was simple, but it cost diagnostic time.

------------------------------------------------------------------------

## Strategy Complexity

The composite signal design went through several iterations. Starting with a plain MA crossover and layering in cluster confirmation and inverse volatility weighting added plausibility but also added debugging difficulty. Each additional term can interact with the others in non-obvious ways, especially after Kalman smoothing.

The Kalman smoothing step is the most complex piece of the pipeline and arguably the least validated. The local-level model assumption (random-walk state) is convenient but not strongly motivated. An EMA-based alternative would have been simpler, more interpretable, and worth testing as a baseline before introducing the Kalman filter. The decision to jump to Kalman smoothing was driven partly by theoretical appeal rather than a demonstrated need.

------------------------------------------------------------------------

## Clustering Utility

Hierarchical clustering was chosen over K-means for its determinism (no random initialisation). This was the right call for reproducibility. However, the clustering's practical value is limited by the small universe size. With eleven sectors, K = 3 produces clusters of roughly three to four sectors each, and the cluster confirmation score does not vary much. In a larger universe (e.g., individual stocks), clustering would carry more information content.

------------------------------------------------------------------------

## Out-of-Sample Evaluation

The most significant methodological limitation is the train/test split approach. The baseline parameters (window = 126, dW = 0.01, K = 3) were selected after observing training performance, introducing implicit data leakage. A walk-forward backtest, where parameters are re-estimated on a rolling basis and signals are generated one period at a time using only past data, would produce a much more credible out-of-sample performance estimate.

The testing period is also too short (14 months) for statistically meaningful conclusions. Extending the dataset and running a rolling-window evaluation would be the correct next step before drawing any real conclusions about edge persistence.

------------------------------------------------------------------------

## Parallel Computation

The grid search parallelisation required more debugging than anticipated. Workers receiving serialised R function closures that retained large enclosing environments caused silent all-NA failures. The fix, creating clean function environments and explicitly exporting only the required objects, worked but added non-trivial complexity. For a 940-combination grid, the sequential implementation would have been simpler and only marginally slower. The engineering overhead of correct parallelisation is worth paying at larger grid sizes, but for smaller searches, simplicity wins.

------------------------------------------------------------------------

# Conclusion

This project implemented a momentum-based sector rotation strategy using the SPDR sector ETFs as the trading universe. The core approach of combining an MA crossover signal, Kalman-smoothed rolling features, hierarchical sector clustering, and inverse-volatility weighting is grounded in a reasonable reading of the momentum literature (Jegadeesh & Titman, 1993; Moskowitz & Grinblatt, 1999; Asness, Moskowitz & Pedersen, 2013), and the pipeline is well-structured and fully reproducible.

Training period performance is negative: the strategy lost money on the data it was designed on. Testing period performance was similarly poor. The grid search over 72 parameter combinations failed to identify a configuration that generates consistent positive in-sample returns, which means the problem is structural rather than one of misconfiguration.

The academic literature provides solid support for sector-level momentum as a concept (Moskowitz & Grinblatt, 1999; Asness, Moskowitz & Pedersen, 2013), but implementing that concept as a working strategy clearly requires more than an MA crossover and hierarchical clustering. The composite signal may be too noisy, the monthly rebalancing frequency may be too slow, the cluster confirmation term may be adding noise rather than signal, or the Kalman smoothing may be damping meaningful variation. Diagnosing which would require systematic ablation testing (running versions of the pipeline with each component removed), which was not done here.

The pipeline itself is structurally sound: the backtest avoids obvious look-ahead bias, the dollar-neutral construction is implemented correctly, and the evaluation is thorough. The engineering is fine. The trading signal is not. That distinction matters for what comes next: the code does not need to be rewritten, but the signal logic does.

------------------------------------------------------------------------

# Appendix: Performance Statistics {.appendix}

```{r, "Appendix Stats Compute", include = F}
# Align strategy and SPY returns by date, then compute all metrics.
period_stats <- function(backtest_result, price_df) {
  strat <- backtest_result$strategy_returns_df %>%
    dplyr::select(date, ret)

  combined <- price_df %>%
    dplyr::filter(symbol == "SPY") %>%
    dplyr::select(date, spy_ret = ClCl) %>%
    dplyr::inner_join(strat, by = "date") %>%
    dplyr::arrange(date)

  rets     <- combined$ret
  spy_rets <- combined$spy_ret
  n        <- nrow(combined)

  cr      <- cumprod(1 + rets)
  cum_ret <- tail(cr, 1) - 1
  spy_cum <- prod(1 + spy_rets) - 1
  ann_ret <- (1 + cum_ret)^(252 / n) - 1
  spy_ann <- (1 + spy_cum)^(252 / n) - 1
  ann_vol <- sd(rets) * sqrt(252)
  sharpe  <- sqrt(252) * mean(rets) / sd(rets)
  mdd     <- min((cr - cummax(cr)) / cummax(cr))
  calmar  <- ann_ret / abs(mdd)
  gains   <- sum(pmax(rets,      0))
  losses  <- sum(pmax(-rets,     0))
  omega   <- if (losses == 0) Inf else gains / losses
  win_rt  <- mean(rets > 0)
  beta    <- cov(rets, spy_rets) / var(spy_rets)
  alpha   <- ann_ret - beta * spy_ann

  monthly <- backtest_result$strategy_returns_df %>%
    dplyr::mutate(.ym = format(date, "%Y-%m")) %>%
    dplyr::group_by(.ym) %>%
    dplyr::summarise(m = prod(1 + ret) - 1, .groups = "drop") %>%
    dplyr::pull(m)

  n_rebal  <- length(unique(backtest_result$weights_df$date))
  in_dd    <- cr < cummax(cr)
  rle_dd   <- rle(in_dd)
  max_dd_d <- if (any(rle_dd$values)) max(rle_dd$lengths[rle_dd$values]) else 0L

  list(cum_ret = cum_ret, spy_cum = spy_cum, excess_cum = cum_ret - spy_cum,
       ann_ret = ann_ret, spy_ann = spy_ann, ann_vol = ann_vol,
       sharpe = sharpe, mdd = mdd, max_dd_d = max_dd_d,
       calmar = calmar, omega = omega, win_rt = win_rt,
       beta = beta, alpha = alpha, avg_mo = mean(monthly), n_rebal = n_rebal)
}

tr <- period_stats(baseline_train, train_df)
te <- period_stats(baseline_test,  test_df)

appendix_stats <- tibble::tibble(
  Group    = c(
    rep("Returns",         5),
    rep("Risk",            3),
    rep("Risk-Adjusted",   4),
    rep("Market Exposure", 2),
    rep("Activity",        2)
  ),
  Metric   = c(
    "Cumulative Return", "SPY Cumulative Return", "Excess vs SPY (Cumulative)",
    "Annualised Return (CAGR)", "SPY Annualised Return",
    "Annualised Volatility", "Maximum Drawdown", "Longest Drawdown (Days)",
    "Sharpe Ratio", "Calmar Ratio", "Omega Ratio", "Daily Win Rate",
    "Beta to SPY", "Annualised Alpha (Jensen's)",
    "Avg Monthly Return", "Number of Rebalances"
  ),
  Training = c(
    tr$cum_ret, tr$spy_cum, tr$excess_cum,
    tr$ann_ret, tr$spy_ann,
    tr$ann_vol, tr$mdd, tr$max_dd_d,
    tr$sharpe, tr$calmar, tr$omega, tr$win_rt,
    tr$beta, tr$alpha,
    tr$avg_mo, tr$n_rebal
  ),
  Testing  = c(
    te$cum_ret, te$spy_cum, te$excess_cum,
    te$ann_ret, te$spy_ann,
    te$ann_vol, te$mdd, te$max_dd_d,
    te$sharpe, te$calmar, te$omega, te$win_rt,
    te$beta, te$alpha,
    te$avg_mo, te$n_rebal
  ),
  # Format type used by gt below; hidden from rendered output
  fmt = c(
    rep("pct", 6), "pct", "int",
    rep("num", 3), "pct",
    rep("num", 2),
    "pct", "int"
  )
)
```

```{r, "Appendix Table - Out", echo = F, warning = F, message = F}
appendix_stats %>%
  gt::gt(groupname_col = "Group") %>%
  gt::tab_header(
    title    = "Strategy Performance Summary",
    subtitle = "Baseline parameters: window = 126, dW = 0.01, K = 3"
  ) %>%
  gt::fmt_percent(
    rows     = fmt == "pct",
    columns  = c(Training, Testing),
    decimals = 2
  ) %>%
  gt::fmt_number(
    rows     = fmt == "num",
    columns  = c(Training, Testing),
    decimals = 2
  ) %>%
  gt::fmt_integer(
    rows    = fmt == "int",
    columns = c(Training, Testing)
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "#c0392b", weight = "bold"),
    locations = gt::cells_body(
      columns = Training,
      rows    = Training < 0 & fmt %in% c("pct", "num")
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_text(color = "#c0392b", weight = "bold"),
    locations = gt::cells_body(
      columns = Testing,
      rows    = Testing < 0 & fmt %in% c("pct", "num")
    )
  ) %>%
  gt::tab_style(
    style     = gt::cell_fill(color = "#f9f9f9"),
    locations = gt::cells_row_groups()
  ) %>%
  gt::cols_hide(columns = fmt) %>%
  gt::cols_label(
    Metric   = gt::md("**Metric**"),
    Training = gt::md("**Training**  \n*(2000–2024)*"),
    Testing  = gt::md("**Testing**  \n*(2025–2026)*")
  ) %>%
  gt::opt_table_font(font = gt::google_font("IBM Plex Sans")) %>%
  gt::tab_options(
    table.width             = gt::pct(85),
    row_group.font.weight   = "bold",
    column_labels.font.size = 13
  )
```

### Metric Definitions

The following metrics are included beyond the basic cumulative return and Sharpe ratio, which are standard. Each captures a different aspect of strategy quality.

**Annualised Return (CAGR).** Compound annual growth rate: the constant annual return that would produce the same total return over the period. More meaningful than cumulative return when comparing periods of very different lengths (24 years training vs 14 months testing).

**Excess vs SPY (Cumulative).** Strategy cumulative return minus SPY cumulative return over the same date range. The most direct answer to whether the active strategy added value over a passive benchmark.

**Maximum Drawdown.** The worst peak-to-trough decline in equity. Answers the question: what is the most an investor would have lost if they entered at the worst time and held until the worst exit? An important complement to Sharpe since two strategies can have identical Sharpe ratios with very different drawdown profiles.

**Calmar Ratio.** Annualised return divided by the absolute maximum drawdown. A risk-adjusted performance measure that is more sensitive to tail losses than Sharpe. A Calmar above 1 is generally considered strong; below 0 means the strategy lost money.

**Omega Ratio.** The probability-weighted ratio of gains to losses above a threshold (here, zero). Unlike Sharpe, it does not assume returns are normally distributed and captures skewness and fat tails. An Omega of 1.0 means gains and losses are balanced; above 1 is net positive, below 1 is net negative.

**Daily Win Rate.** Fraction of trading days with a positive return. Does not capture magnitude, but measures how often the strategy is on the right side. Useful for understanding psychological consistency — a strategy with a 40% win rate can still be profitable if winners are much larger than losers.

**Beta to SPY.** The slope of the regression of strategy daily returns on SPY daily returns. For a correctly implemented dollar-neutral strategy, this should be close to zero. A beta significantly different from zero indicates residual market exposure and that the portfolio is not truly market-neutral.

**Annualised Alpha (Jensen's).** The return unexplained by market beta, annualised. Calculated as annualised strategy return minus beta times annualised SPY return. Represents the pure value-add of the active bets, independent of market direction.

**Longest Drawdown (Days).** The number of consecutive trading days the strategy spent below its prior equity peak. Complements maximum drawdown by measuring duration rather than depth. A long drawdown period is painful even if the ultimate depth is modest.

```{r, "Data Viewer", eval = F, include = F}
# =============================================================================
# DATA VIEWER — not rendered. Run interactively to inspect each pipeline step.
# =============================================================================

# ---- Step 1: Raw Price Data -------------------------------------------------
head(raw_data, 20)
dplyr::glimpse(raw_data)

# ---- Step 2: Train / Test Split ---------------------------------------------
cat("Train rows:", nrow(train_data), "| Test rows:", nrow(test_data), "\n")
head(train_data)
head(test_data)

# ---- Step 3: Returns Added --------------------------------------------------
# OpCl = open-to-close, ClCl = close-to-close, ClOp = close-to-open (overnight)
train_df %>%
  dplyr::select(symbol, date, open, close, OpCl, ClCl, ClOp) %>%
  head(20)

# ---- Step 4: Rolling Stats — Pre-Kalman ------------------------------------
pre_kalman <- train_df %>% add_rollers(window = 126)

pre_kalman %>%
  dplyr::filter(symbol == "XLE") %>%
  dplyr::select(symbol, date, ClCl,
                mean_32d, mean_126d,
                vol_126d, sharpe_126d,
                momentum_126d, corr_spy_126d) %>%
  head(20)

# ---- Step 5: After Kalman Smoothing — Before vs After -----------------------
# Compare a single feature for one symbol to see how much smoothing is applied.
dplyr::bind_cols(
  pre_kalman %>%
    dplyr::filter(symbol == "XLE") %>%
    dplyr::select(date, pre_momentum  = momentum_126d,
                        pre_sharpe    = sharpe_126d,
                        pre_vol       = vol_126d),
  baseline_features %>%
    dplyr::filter(symbol == "XLE") %>%
    dplyr::select(post_momentum = momentum_126d,
                  post_sharpe   = sharpe_126d,
                  post_vol      = vol_126d)
) %>% head(30)

# ---- Step 6: Monthly Snapshots ----------------------------------------------
train_snapshots %>%
  dplyr::select(symbol, date,
                momentum_126d, vol_126d,
                sharpe_126d, corr_spy_126d) %>%
  head(30)

cat("Rebalancing dates:", length(unique(train_snapshots$date)), "\n")

# ---- Step 7: K Evaluation ---------------------------------------------------
k_eval$metrics

# ---- Step 8: Cluster Assignments at a Sample Date --------------------------
sample_rebal    <- max(train_snapshots$date)
sample_clusters <- cluster_sectors(train_snapshots, k = 3,
                                   rebalance_date = sample_rebal)

sample_clusters %>% dplyr::arrange(cluster)

# ---- Step 9: Signals at Sample Date ----------------------------------------
feature_cols_viewer <- c("momentum_126d", "vol_126d",
                         "sharpe_126d",   "corr_spy_126d")

sample_signals <- generate_signals(
  snapshot_dat        = train_snapshots %>% dplyr::filter(date == sample_rebal),
  cluster_assignments = sample_clusters,
  window              = 126
)

sample_signals %>% dplyr::arrange(desc(composite))

# ---- Step 10: Weights at Sample Date ----------------------------------------
sample_weights <- construct_weights(sample_signals)

sample_weights %>% dplyr::arrange(desc(weight))

# ---- Step 11: Portfolio — Daily Weights × Sector Returns --------------------
head(baseline_train$portfolio_df, 30)

# Spot-check: weights should sum near 0 on any given day (dollar-neutral)
baseline_train$portfolio_df %>%
  dplyr::group_by(date) %>%
  dplyr::summarise(net_weight = sum(weight, na.rm = TRUE), .groups = "drop") %>%
  dplyr::summarise(mean_net = mean(net_weight), max_abs_net = max(abs(net_weight)))

# ---- Step 12: Strategy Returns ----------------------------------------------
head(baseline_train$strategy_returns_df, 20)

cat(sprintf(
  "Sharpe: %.2f | CumReturn: %.1f%% | MaxDD: %.1f%%\n",
  baseline_train$sharpe,
  baseline_train$cum_return * 100,
  baseline_train$max_drawdown * 100
))
```
